Restrictive Clearance Comparison

### 2025-02-24 data-raw/RestrictiveClearance_comparison.Rmd
Adjusted fraction unbound comparison. Began moving some data wrangling code to beginning of document RestrictiveClearance_comparison.Rmd

@@

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Kd_estimates_initialsummary.png"),
       bg = "white",
       height = 4, width = 9
)
```

@@

target_chem_lipinski %>%
  group_by(Lipinski) %>%
  count(`Mass <= 500 Da`, `H-donors < 5`, `H-acceptors < 10`, `logP < 5`,
        name = "Count") %>%
  ggplot(aes(
    axis3 = `Mass <= 500 Da`,
    axis4 = `logP < 5`,
    axis2 = `H-donors < 5`,
    axis1 = `H-acceptors < 10`,
    y = Count
  )) +
  geom_alluvium(aes(fill = Lipinski)) +
  geom_stratum() +
  scale_x_discrete(limits = c( 
    "H-acceptors < 10",
    "H-donors < 5",
    "<= 500 Da",
    "logP < 5"
  ),
  position = "top",
  expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  labs(title = "Overwhelmingly, these chemicals passes Lipinksi's Rules of 5",
       fill = "Passes Lipinksi's Rules?") +
  scale_fill_manual(values = c("grey20", "green4")) +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold", color = "black", size = 10),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Alluvium_Lipinski.png"),
       bg = "white",
       height = 5, width = 6
)

# Molecular descriptors in bins
bin_size <- 2
logP_plot <- target_chem_summary %>%
  distinct(dtxsid, logkow) %>% 
  left_join(fup_df, by = join_by(dtxsid == Chemical)) %>%
  mutate(logP_bin = factor(logkow %/% bin_size * bin_size)) %>%
  ggplot(aes(x = logP_bin, y = Funbound.plasma)) +
  labs(x = "logKow", y = "fup") +
  ggbeeswarm::geom_beeswarm() +
  theme_bw()

@@

# LL from pre-calculated httk predictions
restrictive_ll <- all_httk_preds %>% 
  rename(Conc_est = httk_Preds_restrictive,
         Time_trans = Time) %>%
  select(!httk_Preds_nonrestrictive) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!observations) %>%
  mutate(model = "httk_3comp2_restrictive",
         AIC = -2 * log_likelihood)

nonrestrictive_ll <- all_httk_preds %>% 
  rename(Conc_est = httk_Preds_nonrestrictive,
         Time_trans = Time) %>%
  select(!httk_Preds_restrictive) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!observations) %>%
  mutate(model = "httk_3comp2_nonrestrictive",
         AIC = -2 * log_likelihood) # Simple formula when there are no optimized params

httk_AIC <- bind_rows(restrictive_ll, nonrestrictive_ll)

httk_AIC %>%
  distinct(Chemical, Species, model, AIC) %>%
  pivot_wider(names_from = model, values_from = AIC) %>%
  rowwise() %>% 
  mutate(minAIC = min(c_across(starts_with("httk_")), na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(across(starts_with("httk_"), \(x) {
    signif(exp((minAIC - x)/2) * 100, 4)
  })) %>%
  mutate(across(starts_with("httk_"), \(x) {
    ifelse(x < 1E-6, 0, x)
  })) %>% View()

@@

```{r fup_comparison}
fup_df <- this_param_df %>%
  distinct(Chemical, Species, forced_human_values,
           Funbound.plasma, Funbound.plasma.adjustment, 
           Funbound.plasma.dist) %>%
  mutate(
    Funbound.plasma.median = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "^\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.lower95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.upper95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d$"
    ))
    )

>>>>

# Fraction unbound data for target chemicals
fup_df <- this_param_df %>%
  distinct(Chemical, Species, forced_human_values,
           Funbound.plasma, Funbound.plasma.adjustment, 
           Funbound.plasma.dist) %>%
  mutate(
    Funbound.plasma.median = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "^\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.lower95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.upper95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d$"
    ))
    )

human_chems <- httk::get_cheminfo(info = "DTXSID",
                                species = "human",
                                model = "3compartment2",
                                median.only = TRUE,
                                physchem.exclude = TRUE,
                                suppress.messages = TRUE)
human_chems <- intersect(human_chems, new.raw.httk.data$DTXSID)

all_hum_rat_fups <- chem.physical_and_invitro.data %>%
  filter(DTXSID %in% human_chems) %>%
  select(DTXSID, Human.Funbound.plasma, Rat.Funbound.plasma) %>%
  mutate(Rat.Funbound.plasma = as.numeric(str_extract(Rat.Funbound.plasma,
                                                      "^([^,]+)")),
         Human.Funbound.plasma = as.numeric(str_extract(Human.Funbound.plasma,
                                                        "^([^,]+)"))
         )

# Get restrictive and non-restrictive parameters
human_chems_fup <- data.frame(
  Chemical = human_chems,
  fup_restrictive = chem.physical_and_invitro.data$Human.Funbound.plasma[],
                              fup_nonrestrictive = params_nr)
```

@@

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Kd_estimates_initialsummary.png"),
       bg = "white",
       height = 4, width = 9
)

kd_v_fup_plot <- new.raw.httk.data %>%
  inner_join(all_hum_rat_fups) %>%
  pivot_longer(cols = Human.Funbound.plasma:Rat.Funbound.plasma,
               names_to = "Species", values_to = "httk_fup_med") %>%
  ggplot(aes(x = Affinity.Kd.Med, y = httk_fup_med, color = Species)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("green4", "purple2"),
                     labels = c("Human", "Rat")) +
  scale_x_log10(labels = scales::label_log(digits = 2)) +
  labs(x = "Binding affinity ratio", y = "Fraction unbound\nin plasma") +
  theme_bw() +
  theme(legend.position = "top",
        legend.text = element_text(size = 10),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Kd_estimates_vs_fup_multispecies.png"),
       bg = "white",
       height = 3, width = 5
)

@@

lipinski_plot <- target_chem_lipinski %>%
  group_by(Lipinski) %>%
  count(`Mass <= 500 Da`, `H-donors < 5`, `H-acceptors < 10`, `logP < 5`,
        name = "Count") %>%
  ggplot(aes(
    axis3 = `Mass <= 500 Da`,
    axis4 = `logP < 5`,
    axis2 = `H-donors < 5`,
    axis1 = `H-acceptors < 10`,
    y = Count
  )) +
  geom_alluvium(aes(fill = Lipinski)) +
  geom_stratum() +
  scale_x_discrete(limits = c( 
    "H-acceptors < 10",
    "H-donors < 5",
    "<= 500 Da",
    "logP < 5"
  ),
  position = "top",
  expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            size = 2) +
  labs(title = "Target chemicals passing Lipinksi's Rules of 5",
       fill = "Passes Lipinksi's Rules?") +
  scale_fill_manual(values = c("grey20", "green4")) +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold", color = "black", size = 10),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

lipinski_plot
ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Alluvium_Lipinski.png"),
       bg = "white",
       height = 4, width = 5
)

chem_rell_label <- total_rel_evaluation_df %>% distinct(Chemical, ll_label) %>%
  filter(ll_label %in% c("Restrictive", "Nonrestrictive"))

# Molecular descriptors in bins
bin_size <- 10
logP_plot <- target_chem_summary %>%
  inner_join(chem_rell_label, by = join_by(dtxsid == Chemical)) %>%
  distinct(dtxsid, logkow) %>% 
  left_join(fup_df, by = join_by(dtxsid == Chemical)) %>%
  mutate(logP_bin = factor(logkow %/% bin_size * bin_size)) %>%
  ggplot(aes(x = logP_bin, y = Funbound.plasma)) +
  labs(x = "logKow", y = "fup") +
  ggbeeswarm::geom_beeswarm() +
  theme_bw()

target_chem_summary %>%
  inner_join(chem_rell_label, by = join_by(dtxsid == Chemical)) %>%
  distinct(dtxsid, logkow, TPSA, Rot_Bonds, ll_label) %>% 
  pivot_longer(cols = c(logkow, TPSA),
               names_to = "Parameter", values_to = "Value") %>%
  left_join(fup_df, by = join_by(dtxsid == Chemical)) %>%
  ggplot(aes(x = Value, y = Funbound.plasma)) +
  labs(x = "", y = "fup") +
  facet_grid(cols = vars(Parameter),
             rows = vars(ll_label),
             scales = "free_x") +
  geom_point() +
  theme_bw()

@@

# LL from pre-calculated httk predictions
restrictive_ll <- all_httk_preds %>% 
  rename(Conc_est = httk_Preds_restrictive,
         Time_trans = Time) %>%
  select(!httk_Preds_nonrestrictive) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!c(observations, coefs_vector, Time.Units, Time_trans.Units)) %>%
  mutate(model = "model_3comp2_restrictive",
         AIC = -2 * log_likelihood)

nonrestrictive_ll <- all_httk_preds %>% 
  rename(Conc_est = httk_Preds_nonrestrictive,
         Time_trans = Time) %>%
  select(!httk_Preds_restrictive) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!c(observations, coefs_vector, Time.Units, Time_trans.Units)) %>%
  mutate(model = "model_3comp2_nonrestrictive",
         AIC = -2 * log_likelihood) # Simple formula when there are no optimized params

httk_AIC <- bind_rows(restrictive_ll,
                      nonrestrictive_ll,
                      AIC(this_pk) %>% filter(model == "model_flat") %>%
                        filter_targets())

httk_rel_ll <- httk_AIC %>%
  distinct(Chemical, Species, model, AIC) %>%
  pivot_wider(names_from = model, values_from = AIC) %>%
  rowwise() %>% 
  mutate(minAIC = min(c_across(starts_with("model_")), na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(across(starts_with("model_"), \(x) {
    signif(exp((minAIC - x)/2) * 100, 4)
  })) %>%
  mutate(across(starts_with("model_"), \(x) {
    ifelse(x < 1E-6, 0, x)
  })) %>%
  mutate(ll_label = factor(
    case_when(model_flat > 95 ~ "Undetermined",
              model_3comp2_nonrestrictive > 95 &
                model_3comp2_restrictive > 95 ~ "Undetermined (non-flat)",
              model_3comp2_nonrestrictive > 95 ~ "Nonrestrictive",
              model_3comp2_restrictive > 95 ~ "Restrictive")
  ))

httk_rel_ll %>%
  count(model_flat > 95, model_3comp2_nonrestrictive > 95,
        model_3comp2_restrictive > 95)

httk_rel_ll %>%
  select(!minAIC) %>%
  mutate(Chemical = forcats::fct_reorder2(Chemical,
                                          model_3comp2_nonrestrictive,
                                          model_flat,
                                          .fun = mean)) %>%
  pivot_longer(cols = starts_with("model_"),
               names_to = "model", values_to = "relative_likelihood") %>%
  ggplot(aes(x = Chemical, y = model)) +
  geom_tile(aes(fill = relative_likelihood)) +
  scale_fill_viridis_b(option = "cividis", breaks = c(0, 50, 80, 100)) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank())

total_rel_evaluation_df <- httk_rel_ll %>%
  inner_join(this_evaluated_df) %>%
  inner_join(this_param_df %>% distinct(Chemical, forced_human_values)) %>%
  distinct()

total_rel_evaluation_df %>%
  distinct(Chemical, Species, Css.tkstats, Css_httk_nr, Css_httk_r, Css_clear,
         ll_label) %>%
  filter(ll_label %in% c("Nonrestrictive", "Restrictive", "Undetermined")) %>%
  ggplot(aes(x = forcats::fct_reorder(Chemical, Css_httk_r))) +
  geom_pointrange(aes(y = Css.tkstats, ymin = Css_httk_nr, ymax = Css_httk_r),
                  size = 0.1) +
  facet_grid(cols = vars(ll_label), space = "free_x", scales = "free_x") +
  scale_y_log10() +
  labs(x = 'Chemicals', y = "Css") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_blank())

total_rel_evaluation_df %>%
  distinct(Chemical, Species, Css_httk_r,
           halflife.tkstats, restrictive_halflife,
           nonrestrictive_halflife, halflife_clear,
           ll_label) %>%
    filter(ll_label %in% c("Nonrestrictive", "Restrictive", "Undetermined")) %>%
  ggplot(aes(x = forcats::fct_reorder(Chemical, restrictive_halflife))) +
  geom_pointrange(aes(y = halflife.tkstats, 
                      ymin = nonrestrictive_halflife,
                      ymax = restrictive_halflife),
                  size = 0.1) +
  facet_grid(cols = vars(ll_label), space = "free_x", scales = "free_x") +
  scale_y_log10() +
  labs(x = "Chemicals", y = "halflife") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        axis.text.y = element_text(color = "black", size = 10),
        axis.title = element_text(size = 14))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "halflife_fullChart.png"),
       bg = "white",
       height = 3, width = 12
)
```

@@

```{r fup_comparison}
